---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# creditools

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of `creditools` is to put the computational power of an entire risk analytics team into a single, scalable R package. It provides a flexible framework for mathematically simulating and optimizing credit policies. 

Instead of spending weeks writing ad-hoc code to backtest a single credit score, `creditools` allows risk and business analysts to instantly model multi-stage decision funnels (Credit, Anti-fraud, Conversion), simulate the impact of new strategies, and discover the **Optimal Efficient Frontier** between approval volume and default rates.

## Why creditools? (The Business Value)

In modern credit risk management, finding the sweet spot of profitability requires testing endless permutations. `creditools` was designed to answer complex business questions in minutes:

*   **Test N-Scores Simultaneously:** Why validate one challenger score when you can simulate 10 different scores at once? Find out exactly which model yields the best risk-adjusted return.
*   **Find the Optimal Cutoff:** Extract the exact approval rates needed to keep delinquency constant, or find the maximum possible delinquency mitigation while holding your approval volume steady.
*   **Surgical Stress Testing (Swap-ins):** When approving new profiles, delinquency doesn't behave linearly. `creditools` lets you inject custom stress scenarios (e.g., *increase PD by 20% for deciles 9 and 10, 50% for deciles 6-8, and 80% for the rest*).
*   **Preserve Funnel Conversion:** Easily lock empirical conversion and activation rates by stage, ensuring your simulated business volumes reflect real-world borrower behavior.
*   **Massive Scalability:** Under the hood, the engine supports `future` parallel processing. Simulate millions of applicants across hundreds of scenarios effortlessly.

## Installation

You can install the development version of creditools from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("matheuspasche/creditools")
```

## Core Concepts

The package is built around three main ideas:

1.  **`credit_policy()`**: A central object that holds all the configuration for a simulation: column mappings, the sequence of decision stages, and stress scenarios for default simulation.
2.  **`stages`**: A policy is composed of sequential stages, such as `stage_cutoff()` (for score-based decisions) or `stage_rate()` (for probabilistic stages like fraud checks or conversion rates).
3.  **`run_simulation()` / `run_tradeoff_analysis()`**: The simulation engines. `run_simulation()` executes a single, defined policy, while `run_tradeoff_analysis()` is a powerful wrapper that runs dozens or hundreds of simulations to explore the impact of varying parameters.

## A Complete Example

Here is a walk-through of a complex, realistic simulation using a massive analytical base (> 1.5 Million applicants), multiple decision stages (Credit, Fraud, Conversion), parallel processing, and custom stress scenarios.

### 1. Load Package, Parallel Plan and Generate Data

First, we'll generate 1,500,000 applicants mimicking a historical base where applicants have an "old_score" and a new challenger "new_score". We want to evaluate if the `new_score` mitigates default risk while retaining healthy approval and conversion metrics.

```{r example-setup, message=FALSE, warning=FALSE}
library(creditools)
library(dplyr)
library(ggplot2)
library(future)

# Enable parallel processing to handle the > 1.5 Million applicant volume swiftly
future::plan(multisession)

# Generate a massive analytical base (1.5 Million applicants)
sample_data <- generate_sample_data(n_applicants = 1500000, seed = 42)

# Create stratification bands for the new score
sample_data$new_score_decile <- dplyr::ntile(sample_data$new_score, 10)

# Inject simulated Vintages (Time Cohorts) and Categorical dimensions for Advanced Analytics showcase
sample_data$vintage_month <- sample(
  as.Date(c("2023-01-01", "2023-02-01", "2023-03-01", "2023-04-01", "2023-11-01", "2023-12-01")), 
  nrow(sample_data), replace = TRUE
)
sample_data$status <- sample(
  c("Approved", "Denied"), 
  nrow(sample_data), replace = TRUE, prob = c(0.85, 0.15)
)
```

### 2. Define Funnel Stages and Credit Policy

A risk credit policy is rarely a single cutoff. We build a multi-stage funnel where applicants need to sequentially pass through a credit filter, an anti-fraud engine, and finally, a conversion rate probability (e.g. credit seekers are more likely to accept higher rates).

```{r example-policy, message=FALSE, warning=FALSE}
# Stage 1: Credit decision (Approval driven by a Score Cutoff)
credit_stage <- creditools::stage_cutoff(
  name = "credit_decision",
  cutoffs = list(new_score = 600) # This will be dynamically varied later
)

# Stage 2: Anti-fraud model (Flat 95% generic pass rate)
antifraud_stage <- creditools::stage_rate(
  name = "anti_fraud",
  base_rate = 0.95
)

# Stage 3: Conversion rate (Monotonically decreasing with score)
# Worst scores have a robust conversion rate (need credit), best scores have lower rate.
conversion_stage <- creditools::stage_rate(
  name = "conversion",
  base_rate = 0.70, # Baseline, overriding dynamically 
  stress_by_score = list(
    score_col = "new_score",
    rate_at_min = 0.90, 
    rate_at_max = 0.60
  )
)

# Create the full policy object
base_policy <- creditools::credit_policy(
  applicant_id_col = "id",
  score_cols = c("old_score", "new_score"),
  current_approval_col = "approved",
  actual_default_col = "defaulted",
  risk_level_col = "new_score_decile", # Required if we apply stratification in stress test
  simulation_stages = list(
    credit_stage,
    antifraud_stage,
    conversion_stage
  )
)
```

### 3. Inspect a Single Simulation (Granular Audit)

Before running massive loops, you might want to audit exactly what happens to each individual applicant. The `run_simulation()` engine allows you to process a single policy and inspect the exact mathematical flags applied to every single row (e.g., did they pass the cutoff? did they convert? what is their simulated PD? did they default?).

```{r example-single-sim, message=FALSE, warning=FALSE}
# Run a single static pass of the policy over the 1.5M applicants
single_run <- creditools::run_simulation(
  data = sample_data, 
  policy = base_policy,
  quiet = TRUE
)

# Extract the granular data and inspect the first applicant who was a "Swap-In" (approved now, but rejected before)
granular_base <- single_run$data
granular_base %>% 
  dplyr::filter(scenario == "swap_in") %>% 
  dplyr::select(id, old_score, new_score, approved_credit_decision_new, approved_anti_fraud_new, new_approval, simulated_default) %>%
  head()
```

### 4. Run a Massively Parallel Trade-off Analysis

We will define parameters to vary dynamically. We want to test different score cutoffs to see the "swap-in" effects, while also varying the base conversion rate and aggravating the default expectation for newly approved customers (a normal phenomena when exploring unsupervised score bands). 

```{r example-tradeoff, message=FALSE, warning=FALSE}
# Define the simulation grid
vary_params <- list(
  new_score_cutoff = seq(450, 750, by = 50),
  aggravation_factor = c(1.2, 1.5, 1.7) # 20%, 50% and 70% PD Uplift
)

# Run the parallel analysis!
# `run_tradeoff_analysis` intercepts names like `*_base_rate` to dynamically override Funnel Stages
tradeoff_results <- creditools::run_tradeoff_analysis(
  data = sample_data,
  base_policy = base_policy,
  vary_params = vary_params,
  parallel = TRUE, # Using future multisession
  quiet = TRUE
)

head(tradeoff_results)
```

### 5. Visualize the Results

The results provide a data frame mapping out an efficient frontier. This clearly highlights how migrating to `new_score` mitigates default risk depending on the selected cutoff, and showcases the conservative aggravation impacts on business volume.

```{r example-plot, fig.width=9, fig.height=6}
tradeoff_results %>%
  mutate(Stress = paste0("+", round((aggravation_factor - 1) * 100), "% PD Aggravation")) %>%
  ggplot(aes(x = approval_rate, y = default_rate, color = Stress)) +
  geom_line(size = 1.2) +
  geom_point(aes(size = new_score_cutoff), alpha = 0.8) +
  labs(
    title = "Efficient Frontier: Migrating to 'New Score'",
    subtitle = "Trade-off analysis for > 1.5M applicants adjusting for Conversion funnels and PD Aggravations",
    x = "Overall Approval Volume (End of Funnel Rate)", 
    y = "Average Default Rate (%)",
    size = "New Score Cutoff"
  ) +
  scale_x_continuous(labels = scales::percent_format(accuracy=1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy=0.1)) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")
```

## Advanced Analytics: Hard Filters & Risk Based Pricing (RBP)

Beyond point-in-time simulations, `creditools` operates as a robust **Risk Matrix Engine**. Modern credit departments often need to answer: *"Should I replace my primary score with this new provider, or matrix them both into a combined stable Risk Tier?"*

The `find_risk_groups()` heuristic engine solves exactly this. It automatically bins N-scores into intersecting dimensions, calculates empirical risk, merges tiny pockets of poplation (`min_vol_ratio`), and **prunes any groups with high PD volatility across vintages** to guarantee perfectly stable horizontal curves over time (`max_volatility_cv`). Furthermore, you can apply categorical rules (`stage_filter`) prior to evaluation.

Let's simulate a massive dataset with categorical constraints and search for Risk Tiers evaluating the combined predictive power of our 2 scores.

### 1. Hard Filters & Finding Stable Risk Tiers

```{r example-matrix-rating, message=FALSE, warning=FALSE}
# Let's say we only want to evaluate the matrix on "Valid" segments that passed a basic 300 points Cutoff on our internal Score A.
advanced_policy <- credit_policy(
  applicant_id_col = "id",
  score_cols = c("old_score", "new_score"),
  current_approval_col = "approved",
  actual_default_col = "defaulted",
  risk_level_col = "new_score_decile",
  simulation_stages = list(
    # Stage 1: Categorical Filter (Dynamic execution)
    stage_filter(name = "status_filter", condition = "status == 'Approved'"),
    # Stage 2: Pre-Cutoff Screen
    stage_cutoff(name = "baseline", cutoffs = list(old_score = 300))
  )
)

# Pass the data through the hard filters first
filtered_candidates <- run_simulation(sample_data, advanced_policy, quiet = TRUE)$data %>% dplyr::filter(new_approval == TRUE)

# After isolating the approved funnel, we apply the clustering engine!
# We set `oot_date` to blindly validate our hierarchy outside of the train months.
rbp_matrix_results <- creditools::find_risk_groups(
    data = filtered_candidates, 
    score_cols = c("old_score", "new_score"), # Combining both!
    default_col = "defaulted", 
    time_col = "vintage_month",
    bins = 10,                 # Starts as a 10x10 matrix (100 pockets)
    min_vol_ratio = 0.05,      # Groups cannot be smaller than 5%
    max_volatility_cv = 0.20,  # Tolerates maximum 20% of PD Volatility (CV) per Vintage compared to the Average
    oot_date = as.Date("2023-11-01") # Preserves Nov/Dec for Out-Of-Time validation!
)

# How many stable groups did it find?
print(rbp_matrix_results$report)
```

The resulting `rbp_matrix_results$data` comes pre-attached with `risk_rating` (from 1 to N), mapping perfectly to your custom Risk Based Pricing matrix. The results clearly demonstrate if swapping or matrixing scores offers more granular risk differentiation without vintage intersections. This saves weeks of manual SAS/SQL matrix cross-validations!
