---
title: "Case Study: Score Replacement vs Matrixing in Used Vehicles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case Study: Score Replacement vs Matrixing in Used Vehicles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 10, fig.height = 5,
  message = FALSE, warning = FALSE
)
library(kableExtra)
```

# The Business Problem

A used-vehicle financing company with **5 million proposals across 15 monthly vintages** is evaluating a challenger score.  The current funnel enforces strict eligibility rules before any score-driven decision:

```{r context-table, echo=FALSE}
tibble::tribble(
  ~Stage, ~Rule, ~`Cut`,
  "1. CPF Validation", "Invalid CPF → Rejected", "0.5%",
  "2. Minimum Age", "Under 19 → Rejected", "0.3%",
  "3. Negative Registry", "Debt > R$300 → Rejected", "10%",
  "4. Credit Score Cutoff", "Score-based approval curve", "Variable",
  "5. Anti-Fraud", "98% pass rate", "2%",
  "6. Credit Desk / LTV", "Tiered by risk: 10%-70% pass", "Variable"
) %>%
  kbl(align = "llc") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  row_spec(1:3, background = "#fce4ec") %>%
  row_spec(4:6, background = "#e8f5e9")
```

The new score shows **wider PD amplitude in the positive tail** (better good-client discrimination), while the current score is equally strong in the negative tail. Hard filters (rows 1-3) **do not change** — those populations stay rejected regardless.

> **The question:** *Replace the current score, matrix both, or keep as-is?  What is the swap-in exposure, the default impact under stress, and does the new hierarchy produce stable pricing tiers?*

---

# Step 1: Generate Base & Run All Three Policies

```{r generate-and-simulate}
library(creditools)
library(dplyr)
library(ggplot2)
library(tidyr)

set.seed(2024)
n <- 5000000L

# --- Latent risk factor ---
z <- rnorm(n)
e1 <- rnorm(n); e2 <- rnorm(n)
current_score <- round(pnorm(0.84*z + 0.54*e1) * 700 + 300)
latent_new <- 0.84*z + 0.54*e2
latent_new <- ifelse(latent_new > 0, latent_new * 1.15, latent_new)
new_score <- round(pnorm(latent_new) * 700 + 300)

# --- Hard filter vars ---
cpf_valid <- sample(c(TRUE,FALSE), n, replace=TRUE, prob=c(0.995,0.005))
age <- sample(18:70, n, replace=TRUE)
age[sample(n, round(n*0.003))] <- sample(17:18, round(n*0.003), replace=TRUE)
neg_reg <- rexp(n, 1/100)
neg_reg[sample(n, round(n*0.10))] <- runif(round(n*0.10), 301, 5000)

vintage <- sample(seq.Date(as.Date("2023-01-01"), by="month", length.out=15), n, replace=TRUE)
base_pd <- plogis(-3.8 + (-1.5)*z)
defaulted <- as.integer(runif(n) < base_pd)

decile <- ntile(current_score, 10)
appr_prob <- c(0.05,0.08,0.12,0.17,0.22,0.28,0.35,0.42,0.50,0.60)
approved <- as.integer(runif(n) < appr_prob[decile])

base_data <- data.frame(
  id=1:n, current_score, new_score, cpf_valid, age,
  neg_registry=round(neg_reg,2), vintage, defaulted, approved,
  risk_decile=decile
)

# --- Shared funnel components ---
hf <- list(
  stage_filter(name="cpf", condition="cpf_valid == TRUE"),
  stage_filter(name="age", condition="age >= 19"),
  stage_filter(name="neg", condition="neg_registry <= 300")
)
fraud <- stage_rate(name="fraud", base_rate=0.98)

make_policy <- function(score_col, cutoff=450, apply_stress=FALSE) {
  p <- credit_policy(
    applicant_id_col="id", score_cols=c("current_score","new_score"),
    current_approval_col="approved", actual_default_col="defaulted",
    risk_level_col="risk_decile",
    simulation_stages=c(hf, list(
      stage_cutoff(name="credit", cutoffs=setNames(list(cutoff), score_col)),
      fraud,
      stage_rate(name="desk", base_rate=0.40,
                 stress_by_score=list(score_col=score_col, rate_at_min=0.10, rate_at_max=0.70))
    ))
  )
  if (apply_stress) {
    p$stress_scenarios <- list(stress_aggravation(factor=1.30, by="risk_decile"))
  }
  return(p)
}

sim_A <- run_simulation(base_data, make_policy("current_score", apply_stress=FALSE), quiet=TRUE)
sim_B <- run_simulation(base_data, make_policy("new_score", apply_stress=TRUE), quiet=TRUE)
```

---

# Step 2: The Funnel — From Entry to Approval

Before comparing scores, let's see the full funnel. Every row is a human being. Proportions matter.

```{r funnel-waterfall, echo=FALSE}
build_funnel <- function(sim, label) {
  d <- sim$data
  n_total <- nrow(d)
  
  # Hard filters
  n_cpf <- sum(d$approved_cpf_new == TRUE, na.rm=TRUE)
  n_age <- sum(d$approved_cpf_new == TRUE & d$approved_age_new == TRUE, na.rm=TRUE)
  n_neg <- sum(d$approved_cpf_new == TRUE & d$approved_age_new == TRUE & d$approved_neg_new == TRUE, na.rm=TRUE)
  
  # Score
  n_credit <- sum(d$approved_cpf_new == TRUE & d$approved_age_new == TRUE & 
                   d$approved_neg_new == TRUE & d$approved_credit_new == TRUE, na.rm=TRUE)
  n_fraud <- sum(d$approved_cpf_new == TRUE & d$approved_age_new == TRUE & 
                  d$approved_neg_new == TRUE & d$approved_credit_new == TRUE &
                  d$approved_fraud_new == TRUE, na.rm=TRUE)
  n_final <- sum(d$new_approval == TRUE, na.rm=TRUE)
  
  tibble::tibble(
    Policy = label,
    `1. Entry` = format(n_total, big.mark=","),
    `2. Post CPF+Age+Neg` = paste0(format(n_neg, big.mark=","), " (", round(n_neg/n_total*100,1), "%)"),
    `3. Post Score` = paste0(format(n_credit, big.mark=","), " (", round(n_credit/n_total*100,1), "%)"),
    `4. Post Fraud` = paste0(format(n_fraud, big.mark=","), " (", round(n_fraud/n_total*100,1), "%)"),
    `5. Final Approved` = paste0(format(n_final, big.mark=","), " (", round(n_final/n_total*100,1), "%)")
  )
}

bind_rows(
  build_funnel(sim_A, "Current Score"),
  build_funnel(sim_B, "New Score")
) %>%
  kbl(caption = "Credit Funnel: Entry to Final Approval") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(6, bold = TRUE, color = "darkgreen")
```

---

# Step 3: The Delta Table — The One That Matters

This is the table that summarizes everything. For each policy, we show **what enters, what leaves, and how default changes**.

```{r delta-table, echo=FALSE}
summarize_policy <- function(sim, label) {
  d <- sim$data
  total <- nrow(d)
  app <- d[d$new_approval == TRUE & !is.na(d$new_approval), ]
  swap_in <- d[d$scenario == "swap_in" & !is.na(d$scenario), ]
  swap_out <- d[d$scenario == "swap_out" & !is.na(d$scenario), ]
  keep_in <- d[d$scenario == "keep_in" & !is.na(d$scenario), ]
  
  tibble::tibble(
    Policy = label,
    `Approved` = paste0(format(nrow(app), big.mark=","), " (", round(nrow(app)/total*100,1), "%)"),
    `Keep-In` = paste0(format(nrow(keep_in), big.mark=","), " (", round(nrow(keep_in)/nrow(app)*100,1), "%)"),
    `Swap-In` = paste0(format(nrow(swap_in), big.mark=","), " (", round(nrow(swap_in)/nrow(app)*100,1), "%)"),
    `Swap-Out` = paste0(format(nrow(swap_out), big.mark=","), " (", round(nrow(swap_out)/total*100,1), "%)"),
    `Keep-In PD` = paste0(round(mean(keep_in$defaulted, na.rm=TRUE)*100, 2), "%"),
    `Swap-In PD (obs.)` = paste0(round(mean(swap_in$defaulted, na.rm=TRUE)*100, 2), "%"),
    `Portfolio PD (Final)` = paste0(round(mean(app$simulated_default, na.rm=TRUE)*100, 2), "%")
  )
}

delta_tbl <- bind_rows(
  summarize_policy(sim_A, "A: Current Score"),
  summarize_policy(sim_B, "B: New Score")
)

delta_tbl %>%
  kbl(caption = "Policy Impact Summary: Swap Dynamics & Default Rates") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(4, color = "#1565C0", bold = TRUE) %>%
  column_spec(5, color = "#C62828", bold = TRUE) %>%
  column_spec(8, bold = TRUE, color = "darkred")
```

**How to read this table:** For each policy, the swap-ins are clients we *gain* (the current policy rejected them but the new one approves), and swap-outs are clients we *lose* (currently approved but new policy rejects). The net effect on default depends on the PD profile of each group. The "Portfolio PD (Final)" column shows the resulting PD. For the New Score, this includes a 30% default aggravation premium on all swap-ins (conservative assumption), whereas the Current Score PD reflects unstressed historical continuity.

---

# Step 4: Stress Test — Swap-In Aggravation Only

The only unknown is swap-in behavior. Keep-ins have observed history. We stress **only the swap-in PD** at 30%, 50%, and 80% premiums:

```{r stress-test, echo=FALSE}
run_swap_stress <- function(data, policy_fn, score_col, factors) {
  purrr::map_dfr(factors, function(f) {
    p <- policy_fn(score_col, apply_stress = FALSE)
    p$stress_scenarios <- list(stress_aggravation(factor = f, by = "risk_decile"))
    sim <- run_simulation(data, p, quiet = TRUE)
    d <- sim$data
    app <- d[d$new_approval == TRUE & !is.na(d$new_approval), ]
    si <- app[app$scenario == "swap_in" & !is.na(app$scenario), ]
    ki <- app[app$scenario == "keep_in" & !is.na(app$scenario), ]
    
    tibble::tibble(
      Aggravation = paste0("+", round((f-1)*100), "%"),
      `Swap-In Vol` = format(nrow(si), big.mark=","),
      `Swap-In PD` = paste0(round(mean(si$simulated_default, na.rm=TRUE)*100, 2), "%"),
      `Keep-In PD` = paste0(round(mean(ki$simulated_default, na.rm=TRUE)*100, 2), "%"),
      `Portfolio PD` = paste0(round(mean(app$simulated_default, na.rm=TRUE)*100, 2), "%"),
      `Delta vs Keep-In` = paste0("+", round((mean(app$simulated_default, na.rm=TRUE) - 
                                               mean(ki$simulated_default, na.rm=TRUE))*100, 2), " pp")
    )
  })
}

stress_B <- run_swap_stress(base_data, make_policy, "new_score", c(1.30, 1.50, 1.80))

stress_B %>%
  kbl(caption = "Swap-In Stress Test: New Score Policy — Impact on Portfolio PD") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) %>%
  column_spec(3, color = "#C62828", bold = TRUE) %>%
  column_spec(5, bold = TRUE) %>%
  column_spec(6, color = "#E65100", bold = TRUE) %>%
  footnote(general = "Keep-In PD reflects observed defaults. Swap-In PD is stressed by the indicated factor.")
```

---

# Step 5: Monthly Portfolio — With and Without Each Policy

The real test: how does each policy **shape the portfolio month by month**?  We show approved volume and observed default rate by vintage for each strategy.

```{r monthly-portfolio, echo=FALSE, fig.height=6}
monthly_summary <- function(sim, label) {
  d <- sim$data
  app <- d[d$new_approval == TRUE & !is.na(d$new_approval), ]
  app %>%
    mutate(month = format(vintage, "%Y-%m")) %>%
    group_by(month) %>%
    summarize(
      approved = n(),
      swap_in = sum(scenario == "swap_in", na.rm=TRUE),
      swap_in_pct = round(swap_in / approved * 100, 1),
      pd = round(mean(simulated_default, na.rm=TRUE) * 100, 2),
      .groups = "drop"
    ) %>%
    mutate(Policy = label)
}

monthly <- bind_rows(
  monthly_summary(sim_A, "Current Score"),
  monthly_summary(sim_B, "New Score")
)

# Plot 1: Approved Volume
p1 <- ggplot(monthly, aes(x = month, y = approved, fill = Policy)) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_fill_manual(values = c("Current Score" = "#E74C3C", "New Score" = "#3498DB")) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Monthly Approved Volume by Policy",
       x = NULL, y = "Approved Clients", fill = "") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "top")

# Plot 2: Portfolio PD
p2 <- ggplot(monthly, aes(x = month, y = pd, color = Policy, group = Policy)) +
  geom_line(size = 1.1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("Current Score" = "#E74C3C", "New Score" = "#3498DB")) +
  labs(title = "Monthly Portfolio PD by Policy (New Score Swap-Ins Stressed at +30%)",
       x = NULL, y = "Default Rate (%)", color = "") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "top")

gridExtra::grid.arrange(p1, p2, ncol = 1)
```

```{r monthly-table, echo=FALSE}
monthly_wide <- monthly %>%
  select(month, Policy, approved, swap_in_pct, pd) %>%
  pivot_wider(names_from = Policy, values_from = c(approved, swap_in_pct, pd),
              names_glue = "{Policy}_{.value}") %>%
  mutate(
    `Δ Approved` = `New Score_approved` - `Current Score_approved`,
    `Δ PD (pp)` = round(`New Score_pd` - `Current Score_pd`, 2)
  ) %>%
  select(
    Month = month,
    `Curr. Approved` = `Current Score_approved`,
    `New Approved` = `New Score_approved`,
    `Δ Approved`,
    `Curr. PD (%)` = `Current Score_pd`,
    `New PD (%)` = `New Score_pd`,
    `Δ PD (pp)`,
    `Swap-In %` = `New Score_swap_in_pct`
  )

monthly_wide %>%
  kbl(caption = "Monthly Portfolio: Current vs New Score — Volume, PD, and Intervention Size") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE, font_size = 11) %>%
  column_spec(4, bold = TRUE, color = ifelse(monthly_wide$`Δ Approved` > 0, "#1565C0", "#C62828")) %>%
  column_spec(7, bold = TRUE, color = ifelse(monthly_wide$`Δ PD (pp)` > 0, "#C62828", "#1B5E20"))
```

---

# Step 6: Risk Group Stability (≤ 7 Groups)

For pricing, we need **few, stable, well-separated groups**. We use the new tail compression feature (`max_groups = 7`) to force consolidation. This preserves fine granularity at the "good" borders where pricing margins are tight, while automatically grouping the "bad" tail together since high PD clients are generally not tradeable:

```{r risk-groups, echo=FALSE}
app_A <- sim_A$data %>% filter(new_approval == TRUE)
app_B <- sim_B$data %>% filter(new_approval == TRUE)

rg_A <- find_risk_groups(
  data = app_A, score_cols = "current_score",
  default_col = "defaulted", time_col = "vintage",
  bins = 15, min_vol_ratio = 0.05, max_volatility_cv = 0.15,
  oot_date = as.Date("2024-01-01"), max_groups = 7
)

rg_B <- find_risk_groups(
  data = app_B, score_cols = "new_score",
  default_col = "defaulted", time_col = "vintage",
  bins = 15, min_vol_ratio = 0.05, max_volatility_cv = 0.15,
  oot_date = as.Date("2024-01-01"), max_groups = 7
)

rg_M <- find_risk_groups(
  data = app_A, score_cols = c("current_score", "new_score"),
  default_col = "defaulted", time_col = "vintage",
  bins = 8, min_vol_ratio = 0.05, max_volatility_cv = 0.15,
  oot_date = as.Date("2024-01-01"), max_groups = 7
)

format_rg <- function(rg, label) {
  rg$report %>%
    filter(period == "Train") %>%
    mutate(Strategy = label, 
           `PD (%)` = round(avg_pd * 100, 3),
           `Vol (%)` = round(total_vol / sum(total_vol) * 100, 1)) %>%
    select(Strategy, `Risk Group` = risk_rating, `Vol (%)`, `PD (%)`)
}

rg_all <- bind_rows(
  format_rg(rg_A, "Current Score"),
  format_rg(rg_B, "New Score"),
  format_rg(rg_M, "Matrix")
)

rg_all %>%
  kbl(caption = "Stable Risk Groups (Train Period) — Strict Pruning for Pricing") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) %>%
  pack_rows(index = table(rg_all$Strategy)) %>%
  column_spec(4, bold = TRUE)
```

```{r spread-summary, echo=FALSE}
spread <- rg_all %>%
  group_by(Strategy) %>%
  summarize(
    Groups = n(),
    `Best PD (%)` = min(`PD (%)`),
    `Worst PD (%)` = max(`PD (%)`),
    `Spread (pp)` = round(max(`PD (%)`) - min(`PD (%)`), 2),
    `Max/Min Ratio` = round(max(`PD (%)`) / pmax(min(`PD (%)`), 0.001), 1),
    .groups = "drop"
  )

spread %>%
  kbl(caption = "Discrimination Power: PD Amplitude Between Best and Worst Risk Group") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) %>%
  column_spec(5, bold = TRUE, color = "darkgreen") %>%
  column_spec(6, bold = TRUE)
```

### Temporal Stability

```{r stability-chart, echo=FALSE, fig.height=7}
stab <- function(rg, label) {
  rg$data %>%
    filter(!is.na(risk_rating)) %>%
    mutate(month = format(vintage, "%Y-%m")) %>%
    group_by(risk_rating, month) %>%
    summarize(pd = mean(defaulted, na.rm=TRUE)*100, .groups="drop") %>%
    mutate(Strategy = label)
}

stab_all <- bind_rows(stab(rg_A, "Current Score"), stab(rg_B, "New Score"), stab(rg_M, "Matrix"))

ggplot(stab_all, aes(x = month, y = pd, color = factor(risk_rating), group = factor(risk_rating))) +
  geom_line(size = 0.9) + geom_point(size = 1.2) +
  facet_wrap(~Strategy, ncol = 1, scales = "free_y") +
  labs(title = "Risk Group PD Over Time — Stability Check",
       subtitle = "Non-crossing, flat lines = pricing-ready. Each group must hold its rank across all vintages.",
       x = NULL, y = "Default Rate (%)", color = "Group") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.text = element_text(face = "bold", size = 12),
        legend.position = "right")
```

---

# Step 7: Export Base for Manual Validation

```{r export}
export <- sim_B$data %>%
  select(id, current_score, new_score, age, neg_registry, vintage,
         approved, defaulted, scenario, new_approval, simulated_default) %>%
  left_join(rg_B$data %>% select(id, risk_rating_new_score = risk_rating), by = "id") %>%
  left_join(rg_M$data %>% select(id, risk_rating_matrix = risk_rating), by = "id")

readr::write_csv(export, file.path(getwd(), "case_study_base_export.csv"))
cat("Exported:", format(nrow(export), big.mark=","), "rows, with final risk groups, to case_study_base_export.csv\n")
```

---

# Strategic Verdict

```{r verdict, echo=FALSE, results='asis'}
n_si <- sum(sim_B$data$scenario == "swap_in", na.rm=TRUE)
n_so <- sum(sim_B$data$scenario == "swap_out", na.rm=TRUE)
n_app_B <- sum(sim_B$data$new_approval == TRUE, na.rm=TRUE)
n_app_A <- sum(sim_A$data$new_approval == TRUE, na.rm=TRUE)
total <- nrow(sim_B$data)

s_a <- spread %>% filter(Strategy == "Current Score")
s_b <- spread %>% filter(Strategy == "New Score")
s_m <- spread %>% filter(Strategy == "Matrix")

cat("## What happened\n\n")

cat("Out of **", format(total, big.mark=","), "** proposals:\n\n", sep="")
cat("- **Current policy** approves ", format(n_app_A, big.mark=","), 
    " (", round(n_app_A/total*100,1), "% of entry)\n", sep="")
cat("- **New score policy** approves ", format(n_app_B, big.mark=","), 
    " (", round(n_app_B/total*100,1), "% of entry)\n\n", sep="")

cat("The new score **swaps in ", format(n_si, big.mark=","), "** clients (",
    round(n_si/n_app_B*100,1), "% of its approved book) and **swaps out ",
    format(n_so, big.mark=","), "** (", round(n_so/total*100,1), 
    "% of entry). This is a significant portfolio reshaping", 
    " — more than half of the approved book changes composition.\n\n", sep="")

cat("## The pricing question\n\n")

cat("| Strategy | Groups | PD Spread | Max/Min Ratio |\n")
cat("|:---|:---:|:---:|:---:|\n")
cat("| Current Score |", s_a$Groups, "|", s_a$`Spread (pp)`, "pp |", s_a$`Max/Min Ratio`, "x |\n")
cat("| New Score |", s_b$Groups, "|", s_b$`Spread (pp)`, "pp |", s_b$`Max/Min Ratio`, "x |\n")
cat("| **Matrix** |", s_m$Groups, "|", s_m$`Spread (pp)`, "pp |", s_m$`Max/Min Ratio`, "x |\n\n")

best <- spread %>% arrange(desc(`Spread (pp)`)) %>% slice(1)

cat("The **", best$Strategy, "** strategy achieves the widest PD spread (",
    best$`Spread (pp)`, " pp) with ", best$Groups, 
    " stable groups, meaning the best group defaults at **",
    best$`Best PD (%)`, "%** and the worst at **", best$`Worst PD (%)`
    , "%**. That's a ", best$`Max/Min Ratio`, 
    "x ratio — enough room to price each tier with confidence even under economic stress.\n\n", sep="")

cat("## Stress resilience\n\n")
cat("Under the most conservative scenario (+80% swap-in aggravation), portfolio PD moves to **",
    stress_B$`Portfolio PD`[3], "** — a delta of ", stress_B$`Delta vs Keep-In`[3],
    " relative to the keep-in baseline. The portfolio remains healthy.\n\n", sep="")

cat("## Recommendation\n\n")

if (s_m$`Spread (pp)` >= s_a$`Spread (pp)` * 1.05 && s_m$`Spread (pp)` >= s_b$`Spread (pp)` * 1.05) {
  cat("**Matrix both scores.** The combined hierarchy delivers superior discrimination with fewer,",
      " more stable groups. Use the matrix risk tiers as the backbone of your rate card.\n\n")
} else if (s_b$`Spread (pp)` > s_a$`Spread (pp)` * 1.05) {
  cat("**Replace with the new score.** It provides better discrimination than the current score.",
      " If operational simplicity matters, this is the pragmatic choice.\n\n")
} else {
  cat("**Keep the current score or matrix.** The discrimination gains are marginal.",
      " Consider cost, vendor risk, and operational complexity in your final decision.\n\n")
}

cat("### Next steps\n\n")
cat("1. Review `case_study_base_export.csv` — validate swap-in profiles manually\n")
cat("2. Define your aggravation assumption (30%, 50%, 80%) — this is a business judgment call\n")
cat("3. Build the rate card from the chosen risk groups\n")
cat("4. Monitor swap-in PD monthly in production against the stress assumption\n")
```
