---
title: "Case Study: Score Replacement vs Matrixing in Used Vehicles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case Study: Score Replacement vs Matrixing in Used Vehicles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

# Context: The Business Problem

A large used-vehicle financing company processes **~5 million credit applications** spread across **15 monthly vintages**. The credit funnel is complex and sequential:

| Stage | Rule | Rejection Rate |
|:---|:---|:---|
| CPF Validation | Invalid CPF → Rejected | ~0.5% |
| Age | Under 19 → Rejected | ~0.3% |
| Negative Registry | Amount > R$300 → Rejected | ~10% |
| Credit Score (Current) | Best risk tier: 70% pass rate → Worst: 10% | Variable |
| Anti-Fraud | 98% pass rate | ~2% |
| Other Filters (Desk + LTV) | Best tier: 70% → Worst: 10% (by current score) | Variable |

**Observed operational metrics:**

- Final approval rate: ~25%
- Contract conversion rate: 50% of approved
- Portfolio default rate: ~2%

A **new score provider** has been identified. It shows **better tail discrimination** for good risk applicants (wider PD spread in the positive tail), while the **current score performs equally well** in the negative tail. The hard-filter criteria (CPF, age, negative registry) **must not change** regardless of score performance ("keep out" stays "keep out").

**The key strategic question:**

> *Should we replace the current score entirely, or matrix both scores to build a combined risk hierarchy? Under constant approval rate, what happens to default rates if we stress swap-ins by 30%, 50%, and 80%?*

# Step 1: Build the Analytical Base

We generate a synthetic base that closely mirrors the described operation.

```{r generate-base}
library(creditools)
library(dplyr)
library(ggplot2)

set.seed(2024)
n <- 5000000L  # 5 Million clients
n_vintages <- 15

# --- Latent Risk Factor ---
# z: underlying credit quality. Higher z = better client.
z <- rnorm(n)

# --- Scores ---
# Current score: well-calibrated across the range.
# Correlation with latent risk ~ 0.70
e_current <- rnorm(n)
latent_current <- 0.84 * z + 0.54 * e_current
current_score <- round(pnorm(latent_current) * 700 + 300) # Range: ~300-1000

# New score: *better* discrimination in the positive tail (good clients),
# roughly equal in the negative tail.
e_new <- rnorm(n)
latent_new <- 0.84 * z + 0.54 * e_new
# Apply a non-linear stretch to the positive tail
latent_new <- ifelse(latent_new > 0, latent_new * 1.15, latent_new)
new_score <- round(pnorm(latent_new) * 700 + 300) # Range: ~300-1000

# --- Hard Filter Variables ---
cpf_valid <- sample(c(TRUE, FALSE), n, replace = TRUE, prob = c(0.995, 0.005))
age <- sample(18:70, n, replace = TRUE)
# Children under 19 are blocked
age[sample(seq_len(n), round(n * 0.003))] <- sample(17:18, round(n * 0.003), replace = TRUE)
neg_registry <- rexp(n, rate = 1/100)
# ~10% will have > 300
neg_registry[sample(seq_len(n), round(n * 0.10))] <- runif(round(n * 0.10), 301, 5000)

# --- Vintage Assignment (15 months, roughly equal) ---
vintage <- sample(
  seq.Date(as.Date("2023-01-01"), by = "month", length.out = n_vintages),
  n, replace = TRUE
)

# --- Default Outcome ---
# Base PD driven by latent risk, calibrated so portfolio PD ~ 2%
base_pd <- plogis(-3.8 + (-1.5) * z)
defaulted <- as.integer(runif(n) < base_pd)

# --- Current Approval (drives the "keep-in" scenario) ---
# Simulated so final approval ~ 25%
current_score_decile <- ntile(current_score, 10)
# Probability of being approved in the current policy varies by decile
approval_prob_by_decile <- c(0.05, 0.08, 0.12, 0.17, 0.22, 0.28, 0.35, 0.42, 0.50, 0.60)
approved <- as.integer(runif(n) < approval_prob_by_decile[current_score_decile])

# Assemble into a data frame
base_data <- data.frame(
  id = seq_len(n),
  current_score = current_score,
  new_score = new_score,
  cpf_valid = cpf_valid,
  age = age,
  neg_registry = round(neg_registry, 2),
  vintage = vintage,
  defaulted = defaulted,
  approved = approved
)

# Bind the risk decile for stress stratification
base_data$risk_decile <- ntile(base_data$current_score, 10)

cat("Base generated:", format(nrow(base_data), big.mark = ","), "rows\n")
cat("Vintages:", n_vintages, "\n")
cat("Current approval rate:", round(mean(base_data$approved) * 100, 1), "%\n")
cat("Portfolio PD (approved only):", 
    round(mean(base_data$defaulted[base_data$approved == 1]) * 100, 2), "%\n")
```

# Step 2: Build the Multi-Stage Credit Funnel

We recreate the exact funnel described by the business using `creditools`.

```{r build-funnel}
# ---- Define the Hard Filters ----
# These are non-negotiable. Even if the new score says "approve", these rules stay.
filter_cpf <- stage_filter(
  name = "cpf_validation",
  condition = "cpf_valid == TRUE"
)

filter_age <- stage_filter(
  name = "age_check",
  condition = "age >= 19"
)

filter_negativacao <- stage_filter(
  name = "neg_registry",
  condition = "neg_registry <= 300"
)

# ---- Define the Score-Based Credit Decision ----
# This is where the "new_score" cutoff will dynamically vary.
# For the baseline, we use the current score.
credit_stage <- stage_cutoff(
  name = "credit_decision",
  cutoffs = list(current_score = 450)  # Baseline cutoff on current score
)

# ---- Anti-Fraud: 98% pass rate ----
fraud_stage <- stage_rate(
  name = "anti_fraud",
  base_rate = 0.98
)

# ---- Other Filters (Desk/LTV): Monotonic by risk tier ----
# Best decile: 70% pass, worst decile: 10% pass
other_filters <- stage_rate(
  name = "other_approvals",
  base_rate = 0.40,
  stress_by_score = list(
    score_col = "current_score",
    rate_at_min = 0.10,
    rate_at_max = 0.70
  )
)

# ---- Assemble the Full Policy ----
baseline_policy <- credit_policy(
  applicant_id_col = "id",
  score_cols = c("current_score", "new_score"),
  current_approval_col = "approved",
  actual_default_col = "defaulted",
  risk_level_col = "risk_decile",
  simulation_stages = list(
    filter_cpf,
    filter_age,
    filter_negativacao,
    credit_stage,
    fraud_stage,
    other_filters
  ),
  stress_scenarios = list(
    stress_aggravation(factor = 1.30, by = "risk_decile")  # 30% default aggravation on swap-ins
  )
)

# ---- Run the Baseline Simulation ----
baseline_sim <- run_simulation(base_data, baseline_policy, quiet = TRUE)
baseline_summary <- summarize_results(baseline_sim)
print(baseline_summary)
```

# Step 3: Risk Group Discovery (Current Score Only)

Before comparing scores, let's discover how many **stable Homogeneous Groups (GH)** the current score alone can produce.

```{r risk-groups-current}
# Filter only the approved population (keep-ins + swap-ins)
approved_pop <- baseline_sim$data %>%
  filter(new_approval == TRUE)

cat("Approved population:", format(nrow(approved_pop), big.mark = ","), "\n")

# Find Risk Groups using ONLY the current score
rg_current <- find_risk_groups(
  data = approved_pop,
  score_cols = "current_score",
  default_col = "defaulted",
  time_col = "vintage",
  bins = 20,
  min_vol_ratio = 0.05,
  max_volatility_cv = 0.15,
  oot_date = as.Date("2024-01-01")  # Last 2 months as OOT
)

cat("Stable Risk Groups (Current Score):", 
    length(unique(rg_current$report$risk_rating[rg_current$report$period == "Train"])), "\n\n")
print(rg_current$report)
```

# Step 4: Risk Group Discovery (Matrixing Both Scores)

Now let's see if **combining both scores** yields a finer, more stable risk hierarchy.

```{r risk-groups-matrix}
# Find Risk Groups using BOTH scores (2D Matrix)
rg_matrix <- find_risk_groups(
  data = approved_pop,
  score_cols = c("current_score", "new_score"),
  default_col = "defaulted",
  time_col = "vintage",
  bins = 10,            # 10x10 = 100 initial pockets
  min_vol_ratio = 0.05,
  max_volatility_cv = 0.15,
  oot_date = as.Date("2024-01-01")
)

cat("Stable Risk Groups (Matrix):", 
    length(unique(rg_matrix$report$risk_rating[rg_matrix$report$period == "Train"])), "\n\n")
print(rg_matrix$report)
```

# Step 5: Stress Scenarios — Swap-In Aggravation

The critical question: **what happens to PD if we maintain the same approval rate but aggravate swap-in default rates by 30%, 50%, and 80%?**

```{r stress-comparison}
run_stress_scenario <- function(data, policy_base, aggravation_factor) {
  # Clone the policy with a new stress factor
  stressed_policy <- policy_base
  stressed_policy$stress_scenarios <- list(
    stress_aggravation(factor = aggravation_factor, by = "risk_decile")
  )
  
  sim <- run_simulation(data, stressed_policy, quiet = TRUE)
  sim_data <- sim$data
  
  approved <- sim_data[sim_data$new_approval == TRUE & !is.na(sim_data$new_approval), ]
  
  tibble::tibble(
    aggravation = paste0("+", round((aggravation_factor - 1) * 100), "%"),
    aggravation_factor = aggravation_factor,
    total_approved = nrow(approved),
    approval_rate = round(nrow(approved) / nrow(sim_data) * 100, 2),
    swap_in_count = sum(sim_data$scenario == "swap_in", na.rm = TRUE),
    keep_in_count = sum(sim_data$scenario == "keep_in", na.rm = TRUE),
    portfolio_pd = round(mean(approved$simulated_default, na.rm = TRUE) * 100, 3),
    swap_in_pd = round(mean(
      approved$simulated_default[approved$scenario == "swap_in"], na.rm = TRUE
    ) * 100, 3),
    keep_in_pd = round(mean(
      approved$simulated_default[approved$scenario == "keep_in"], na.rm = TRUE
    ) * 100, 3)
  )
}

# Run all three stress scenarios
stress_results <- dplyr::bind_rows(
  run_stress_scenario(base_data, baseline_policy, 1.30),
  run_stress_scenario(base_data, baseline_policy, 1.50),
  run_stress_scenario(base_data, baseline_policy, 1.80)
)

knitr::kable(
  stress_results,
  caption = "Impact of Swap-In Default Aggravation on Portfolio PD",
  col.names = c("Aggravation", "Factor", "Approved", "Appr. Rate %", 
                "Swap-Ins", "Keep-Ins", "Portfolio PD %", 
                "Swap-In PD %", "Keep-In PD %")
)
```

# Step 6: Comparing Risk Hierarchies

Let's visualize the **PD spread** (amplitude between best and worst risk group) for each approach.

```{r compare-hierarchies, fig.height=6}
# Prepare Train-only comparison data
train_current <- rg_current$report %>%
  filter(period == "Train") %>%
  mutate(strategy = "Current Score Only")

train_matrix <- rg_matrix$report %>%
  filter(period == "Train") %>%
  mutate(strategy = "Matrix (Current + New)")

comparison <- bind_rows(train_current, train_matrix) %>%
  group_by(strategy) %>%
  mutate(
    rank = row_number(),
    total_ranks = n()
  ) %>%
  ungroup()

ggplot(comparison, aes(x = rank, y = avg_pd * 100, color = strategy)) +
  geom_line(size = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("Current Score Only" = "#E74C3C", 
                                 "Matrix (Current + New)" = "#2ECC71")) +
  labs(
    title = "PD by Risk Group: Current Score vs Matrix",
    subtitle = "Higher spread = better pricing discrimination",
    x = "Risk Group (Best → Worst)",
    y = "Default Rate (%)",
    color = "Strategy"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

# Step 7: Technical Verdict

```{r verdict}
# Compute key metrics for comparison
spread_current <- diff(range(
  rg_current$report$avg_pd[rg_current$report$period == "Train"]
)) * 100

spread_matrix <- diff(range(
  rg_matrix$report$avg_pd[rg_matrix$report$period == "Train"]
)) * 100

n_groups_current <- length(unique(
  rg_current$report$risk_rating[rg_current$report$period == "Train"]
))

n_groups_matrix <- length(unique(
  rg_matrix$report$risk_rating[rg_matrix$report$period == "Train"]
))

cat("=" |> rep(60) |> paste(collapse = ""), "\n")
cat("                STRATEGIC RECOMMENDATION\n")
cat("=" |> rep(60) |> paste(collapse = ""), "\n\n")

cat("Risk Groups (Current Score Only):", n_groups_current, "\n")
cat("Risk Groups (Matrix):", n_groups_matrix, "\n")
cat("PD Spread - Current:", round(spread_current, 2), "pp\n")
cat("PD Spread - Matrix:", round(spread_matrix, 2), "pp\n")
cat("Spread Gain:", round(spread_matrix - spread_current, 2), "pp\n\n")

if (spread_matrix > spread_current * 1.10) {
  cat(">> RECOMMENDATION: MATRIX BOTH SCORES.\n")
  cat(">> The combined matrix provides", 
      round((spread_matrix / spread_current - 1) * 100, 1), 
      "% more PD amplitude.\n")
  cat(">> This translates to finer Risk-Based Pricing tiers and\n")
  cat(">> better risk-adjusted returns even under conservative\n")
  cat(">> swap-in aggravation scenarios.\n")
} else if (spread_matrix > spread_current) {
  cat(">> RECOMMENDATION: MARGINAL GAIN FROM MATRIXING.\n")
  cat(">> Consider matrixing if the operational cost is low.\n")
  cat(">> Otherwise, replacing the score may simplify governance.\n")
} else {
  cat(">> RECOMMENDATION: REPLACE WITH NEW SCORE.\n")
  cat(">> Matrixing does not add meaningful discrimination.\n")
  cat(">> A simpler single-score policy is preferred.\n")
}

cat("\n")
cat("-" |> rep(60) |> paste(collapse = ""), "\n")
cat("Stress Test Summary (at constant approval rate):\n")
print(stress_results %>% select(aggravation, portfolio_pd, swap_in_pd, keep_in_pd))
```

# Conclusion

This case study demonstrates the **end-to-end capabilities of `creditools`** for a complex real-world credit operation:

1. **Hard Filters** (`stage_filter`): Non-negotiable business rules (CPF, age, negative registry) are enforced before any score-based decision. These populations remain "keep out" regardless of which score is used.

2. **Multi-Stage Funnel** (`credit_policy` + `run_simulation`): The complete decision chain — from hard filters through score cutoffs, fraud checks, and desk approvals — is modeled as a sequential funnel with granular per-applicant tracking.

3. **Risk Grouping** (`find_risk_groups`): Stable, non-crossing risk hierarchies are automatically discovered, pruned for volume, and validated for temporal stability (vintage volatility). The algorithm detects whether matrixing adds genuine discrimination or whether a single score suffices.

4. **Stress Testing** (`stress_aggravation`): Swap-in populations are aggraved at 30%, 50%, and 80% premiums to understand the portfolio's sensitivity to adverse scenarios under constant approval constraints.

5. **Actionable Output**: The final recommendation is data-driven: *matrix or replace*, supported by quantified PD spread gains, stress-test resilience, and OOT validation metrics.
