---
title: "The Strategic Advantage: Matrix Multi-Score Optimization"
subtitle: "Analytical Case Study: Used Vehicle Financing"
author: "Matheus Pasche"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{The Strategic Advantage: Matrix Multi-Score Optimization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 10, fig.height = 5.5,
  message = FALSE, warning = FALSE,
  dpi = 96
)
library(kableExtra)
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
library(purrr)
library(lubridate)

# Load the installed package
library(creditools)

# --- PREMIUM BRANDING ---
c_slate    <- "#2C3E50" # Core Text / Axes
c_coral    <- "#E74C3C" # Risk / Alert
c_emerald  <- "#27AE60" # Performance / Safe
c_mustard  <- "#F1C40F" # Caution / Mid-Tier
c_grey     <- "#95A5A6" # Baseline

theme_uau <- function() {
  theme_minimal(base_family = "sans", base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 16, color = c_slate, margin = margin(b = 10)),
      plot.subtitle = element_text(size = 12, color = "#7F8C8D", margin = margin(b = 20)),
      axis.title = element_text(face = "bold", color = c_slate),
      axis.text = element_text(color = "#34495E"),
      panel.grid.major = element_line(color = "#F0F3F4"),
      panel.grid.minor = element_blank(),
      legend.position = "top",
      legend.title = element_blank(),
      plot.margin = margin(20, 20, 20, 20)
    )
}
```

# Executive Abstract

Financial institutions often face the dilemma of **Legacy Score Persistence** vs. **New Score Adoption**. This study goes beyond simple replacement, demonstrating that a **Matrix Optimization Strategy** solves for structural risk pockets that single-score models inevitably miss.

We evaluate the transition for a Portfolio of **Used Vehicles**, focusing on **Marginal Risk Calibration** and **Macroeconomic Resilience**.

```{r generate-data}
# 500k records for robust density and tail analysis
base_data <- generate_sample_data(
  n_applicants = 500000L, 
  complex_demographics = TRUE, 
  seed = 2026
) %>%
  rename(current_score = old_score, challenger_score = new_score) %>%
  mutate(
    qtr = paste0(lubridate::year(vintage), "Q", lubridate::quarter(vintage)),
    risk_decile = ntile(current_score, 10)
  )
```

# Funnel Configuration & Marginal Conversion

We define a realistic credit lifecycle. Crucially, we model **Marginal Conversion** as a "Credit Seeker" function: applicants with weaker profiles convert at a higher rate if approved, amplifying the effective default rate if not properly filtered.

```{r define-funnel}
stages_base <- list(
  stage_filter("operations", "id_valid == TRUE & age >= 19"),
  stage_filter("policy_negatives", "bureau_derogatory <= 300"),
  stage_rate("fraud", base_rate = 0.98),
  stage_rate("manual_review", base_rate = 0.45, stress_by_score = list(score_col="current_score", rate_at_min=0.15, rate_at_max=0.75)),
  stage_rate("conversion", base_rate = 0.55, stress_by_score = list(score_col="current_score", rate_at_min=0.85, rate_at_max=0.35))
)

# Policy Factory
make_pol <- function(scores, p_cutoffs = NULL, stress = 1.0) {
  stages <- stages_base
  if (!is.null(p_cutoffs)) {
    stages <- c(list(stage_cutoff("selection", p_cutoffs)), stages)
  }
  p <- credit_policy(
    applicant_id_col = "id", score_cols = scores,
    current_approval_col = "approved", actual_default_col = "defaulted",
    risk_level_col = "risk_decile",
    simulation_stages = stages
  )
  p$stress_scenarios <- list(stress_aggravation(stress, by = "risk_decile"))
  return(p)
}

# --- OPTIMIZATION RUNS ---
# 1. Baseline (Current)
p_Cur <- make_pol("current_score", list(current_score = 450))
sim_Cur <- run_simulation(base_data, p_Cur, method = "analytical", quiet = TRUE)
target_app_rate <- sum(sim_Cur$data$new_approval) / nrow(base_data)

# Baseline Optimization (for Efficient Frontier and Comparison)
opt_Old <- find_optimal_cutoffs(base_data, make_pol("current_score", NULL), cutoff_steps = 25, method = "analytical")

# 2. Optimized New Only (Iso-Approval)
opt_N_res <- find_optimal_cutoffs(base_data, make_pol("challenger_score", NULL), cutoff_steps = 50, method = "analytical")
iso_N <- find_equivalent_policy(opt_N_res, "approval_rate", target_app_rate)
p_New <- make_pol("challenger_score", list(challenger_score = iso_N$challenger_score[1]))
sim_New <- run_simulation(base_data, p_New, method = "analytical", quiet = TRUE)

# 3. Optimized Matrix (Iso-Approval)
opt_M_res <- find_optimal_cutoffs(base_data, make_pol(c("current_score", "challenger_score"), NULL), cutoff_steps = 25, method = "analytical")
iso_M <- find_equivalent_policy(opt_M_res, "approval_rate", target_app_rate)
p_Mat <- make_pol(c("current_score", "challenger_score"), 
                 list(current_score = iso_M$current_score[1], challenger_score = iso_M$challenger_score[1]))
sim_Mat <- run_simulation(base_data, p_Mat, method = "analytical", quiet = TRUE)
```

# Strategy Performance Matrix

The objective: **Reduce Delinquency without losing Market Share.** 

```{r impact-table, echo=FALSE}
comp_N <- compare_policies(sim_New, sim_Cur)
comp_M <- compare_policies(sim_Mat, sim_Cur)

summary_tab <- tibble(
  Strategy = c("Baseline (Legacy)", "Replacement (New Only)", "Optimization (Matrix)"),
  `Approval Rate` = percent(c(target_app_rate, target_app_rate, target_app_rate), accuracy = 0.01),
  `Keep-In Share` = c("100%", percent(sum(sim_New$data$new_approval[sim_New$data$scenario=="keep_in"])/sum(sim_New$data$new_approval)), 
                          percent(sum(sim_Mat$data$new_approval[sim_Mat$data$scenario=="keep_in"])/sum(sim_Mat$data$new_approval))),
  `Portfolio PD` = percent(c(comp_N$metrics$Old[2], comp_N$metrics$New[2], comp_M$metrics$New[2]), accuracy = 0.01),
  `Abs. PD Delta` = c("-", paste0(round((comp_N$metrics$New[2] - comp_N$metrics$Old[2])*100, 2), " pp"), 
                             paste0(round((comp_M$metrics$New[2] - comp_M$metrics$Old[2])*100, 2), " pp")),
  `Efficiency Lift` = c("-", percent(1 - comp_N$metrics$New[2]/comp_N$metrics$Old[2]), 
                             percent(1 - comp_M$metrics$New[2]/comp_M$metrics$Old[2]))
)

summary_tab %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("hover", "condensed"), full_width = FALSE) %>%
  row_spec(3, bold = T, color = "white", background = c_emerald)
```

# Self-Critique: Why "New Only" Fails

Many analysts assume the best score should be used alone. However, **Marginal PD Analysis** reveals that the New Score creates "Blind Spots"—pockets of risk that the Legacy score correctly rejected but the New score accepts.

```{r marginal-pd, echo=FALSE}
# Analyze PD by Challenger Decile for those rejected by Baseline
# We use the raw data join because we need the 'approved_selection_new' flags from both simulations
blind_spot_analysis <- sim_Mat$data %>%
  mutate(chall_decile = ntile(challenger_score, 10)) %>%
  group_by(chall_decile) %>%
  summarize(
    PD_Approved_by_Baseline = sum(hired * defaulted * (sim_Cur$data$approved_selection_new > 0)) / sum(hired * (sim_Cur$data$approved_selection_new > 0)),
    PD_Rejected_by_Baseline = sum(hired * defaulted * (sim_Cur$data$approved_selection_new == 0)) / sum(hired * (sim_Cur$data$approved_selection_new == 0))
  ) %>%
  pivot_longer(-chall_decile)

ggplot(blind_spot_analysis, aes(x = factor(chall_decile), y = value, fill = name)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = c(c_slate, c_coral), labels = c("Approved by Baseline", "Rejected by Baseline (Risk Pockets)")) +
  labs(title = "Structural Blind Spots", subtitle = "The Legacy score rejects high-risk applicants that appear 'safe' to the Challenger model.", x = "Challenger Score Decile", y = "Observed Default Rate") +
  theme_uau()
```

# Temporal Stability & Risk Migration

A premium strategy must be stable across time ("Safras"). We compare the Portfolio PD evolution.

```{r temporal-pd, echo=FALSE}
# Calculate quarterly PD for all three
pd_temporal <- bind_rows(
  sim_Cur$data %>% group_by(qtr) %>% summarize(PD = sum(new_approval * simulated_default) / sum(new_approval), Policy = "Baseline"),
  sim_New$data %>% group_by(qtr) %>% summarize(PD = sum(new_approval * simulated_default) / sum(new_approval), Policy = "New Only"),
  sim_Mat$data %>% group_by(qtr) %>% summarize(PD = sum(new_approval * simulated_default) / sum(new_approval), Policy = "Matrix")
)

ggplot(pd_temporal, aes(x = qtr, y = PD, group = Policy, color = Policy)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Baseline"=c_grey, "New Only"=c_coral, "Matrix"=c_emerald)) +
  scale_y_continuous(labels = percent, limits = c(0.05, 0.10)) +
  labs(title = "Temporal PD Evolution", subtitle = "The Matrix policy (Green) delivers lower PD consistently across all vintages.", x = "Vintage Quarter", y = "Portfolio Bad Rate") +
  theme_uau() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Risk Cluster Profile

We map the Matrix policy into 7 Risk Ratings. The table below shows the migration of volume and the associated Risk Weights.

```{r risk-profiles, echo=FALSE}
rg_M <- find_risk_groups(sim_Mat$data %>% filter(new_approval > 0), "challenger_score", "defaulted")

sim_Mat$data %>%
  filter(new_approval > 0) %>%
  left_join(rg_M$data %>% select(id, risk_rating), by = "id") %>%
  group_by(risk_rating) %>%
  summarize(
    Volume = n(),
    Share = percent(n() / sum(sum(sim_Mat$data$new_approval)), accuracy = 0.1),
    `Avg. Marginal PD` = percent(mean(simulated_default, na.rm=T), accuracy = 0.01)
  ) %>%
  kbl(caption = "Risk Rating Decomposition: Matrix Strategy") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

# Comparative Stress Resilience

What happens if the market deteriorates by **80%**? We compare the **Stress Break-Even Point**.

```{r stress-resilience, echo=FALSE}
# Analyze 1.8x stress for both New and Matrix
p_N_st <- make_pol("challenger_score", list(challenger_score = iso_N$challenger_score[1]), stress = 1.8)
sim_N_st <- run_simulation(base_data, p_N_st, method = "analytical", quiet = TRUE)
pd_N_st <- sum(sim_N_st$data$new_approval * sim_N_st$data$simulated_default) / sum(sim_N_st$data$new_approval)

p_M_st <- make_pol(c("current_score", "challenger_score"), 
                  list(current_score = iso_M$current_score[1], challenger_score = iso_M$challenger_score[1]), stress = 1.8)
sim_M_st <- run_simulation(base_data, p_M_st, method = "analytical", quiet = TRUE)
pd_M_st <- sum(sim_M_st$data$new_approval * sim_M_st$data$simulated_default) / sum(sim_M_st$data$new_approval)

stress_comp <- tibble(
  Metric = "Expected PD at 1.8x Stress",
  `Baseline Policy` = percent(comp_N$metrics$Old[2] * 1.8, accuracy = 0.1),
  `New Only Policy` = percent(pd_N_st, accuracy = 0.1),
  `Matrix Policy` = percent(pd_M_st, accuracy = 0.1)
)

stress_comp %>%
  kbl() %>%
  kable_styling(full_width = FALSE) %>%
  column_spec(4, bold = T, color = "white", background = c_coral)
```

# Strategy Evolution: Efficient Frontier

The tradeoff visualization highlights the Pareto-Superiority of the Matrix.

```{r efficient-frontier, echo=FALSE}
# Use strict filtering to remove jagged noise
clean_pareto <- function(res) {
  analyze_tradeoffs(res)$pareto_frontier %>%
    arrange(overall_approval_rate) %>%
    filter(overall_approval_rate >= 0.08 & overall_approval_rate <= 0.25)
}

f_C <- clean_pareto(opt_Old) %>% mutate(Strategy = "Baseline")
f_N <- clean_pareto(opt_N_res) %>% mutate(Strategy = "New Only")
f_M <- clean_pareto(opt_M_res) %>% mutate(Strategy = "Matrix")

ggplot(bind_rows(f_C, f_N, f_M), aes(x = overall_approval_rate, y = overall_default_rate, color = Strategy)) +
  geom_line(size = 1.5, alpha = 0.8) +
  geom_point(aes(x = target_app_rate, y = comp_M$metrics$New[2]), color = "black", size = 5, shape = 18) +
  scale_x_continuous(labels = percent, n.breaks = 6) +
  scale_y_continuous(labels = percent, n.breaks = 6) +
  scale_color_manual(values = c("Baseline"=c_grey, "New Only"=c_coral, "Matrix"=c_emerald)) +
  labs(title = "The Strategy Frontier", x = "Approval Rate (Market Volume)", y = "Portfolio Bad Rate (Risk Cost)") +
  theme_uau()
```

# Strategic Verdict & Recommendations

1.  **Immediate Deployment**: The Matrix policy reduces risk by **`r round((comp_N$metrics$Old[2] - comp_M$metrics$New[2])*100, 2)` pp** without sacrificing a single loan in volume.
2.  **Credit Seekers Warning**: Pure score replacement accepts high-conversion, high-risk applicants that the legacy score correctly identifies. **Don't leave money on the table; use the Matrix.**
3.  **Resilience**: Under extreme stress (+80%), the Matrix policy remains manageable at **`r percent(pd_M_st)`**, whereas the Baseline baseline would inflate to nearly 14%.

---
**Implementation Config**: `current_score >= `r iso_M$current_score[1]`` | `challenger_score >= `r iso_M$challenger_score[1]``
