---
title: "Strategic Decision Engine: Optimizing Used Vehicle Credit Strategy"
subtitle: "Proposed Transition: From Legacy Score to Optimized Matrix Strategy"
author: "Matheus Pasche (Strategic Analyst Portfolio)"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    css: !expr "system.file('rmarkdown/templates/html_vignette/resources/vignette.css', package = 'rmarkdown')"
vignette: >
  %\VignetteIndexEntry{Strategic Credit Decision Engine: Used Vehicle Portfolio Optimization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 4.5,
  out.width = "100%",
  message = FALSE, warning = FALSE,
  cache = FALSE,
  dpi = 130
)
library(creditools)
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
library(purrr)
library(lubridate)
library(kableExtra)

# --- REFINED COLOR PALETTE ---
c_slate    <- "#2C3E50" # Primary Text
c_emerald  <- "#27AE60" # Matrix Strategy
c_ruby     <- "#E74C3C" # High Stress / Replacement
c_orange   <- "#E67E22" # Mid Stress
c_gold     <- "#D4AC0D" # Low Stress
c_sapphire <- "#2980B9" # Legacy Strategy

# Professional Theme
theme_strategic <- function() {
  theme_minimal(base_family = "sans", base_size = 10) +
    theme(
      plot.title = element_text(face = "bold", size = 12, color = c_slate),
      plot.subtitle = element_text(size = 10, color = "#7F8C8D"),
      axis.title = element_text(face = "bold", color = c_slate),
      axis.text = element_text(color = "#34495E"),
      panel.grid.major = element_line(color = "#E5E8E8"),
      panel.grid.minor = element_blank(),
      legend.position = "top",
      legend.justification = "left",
      legend.title = element_blank()
    )
}

fmt_pct <- function(x) percent(x, accuracy = 0.1)
```

# Project Narrative & Strategic Context

In the used vehicle financing sector, the transition from a "Legacy Score" to a "Challenger" model is as much about risk management as it is about predictive power. A pure replacement often introduces "Swap Risk": rejections of historically safe applicants and new exposures without vintage validation.

This report evaluates the optimal path: **Matrix Strategy**. We focus on maintaining our historical approval levels while systematically reducing the Expected PD under stressful macroeconomic conditions (1.5x aggravation).

---

# 1. Baseline Diagnosis: Portfolio DNA

We analyze a simulated portfolio of **500,000 applicants**. Our starting point is the "Ground Truth"—the historically realized performance ("Ele é ele").

```{r data-prep}
# Restoration of 500k Applicants for full statistical significance
raw_data <- generate_sample_data(
  n_applicants = 500000L, 
  complex_demographics = TRUE, 
  seed = 42 
) %>%
  rename(legacy_score = old_score, challenger_score = new_score) %>%
  mutate(
    legacy_decile = ntile(legacy_score, 10),
    challenger_decile = ntile(challenger_score, 10),
    vintage_month = floor_date(as.Date(vintage), "month")
  )

# --- BASELINE SNAPSHOT ---
h_app <- raw_data %>% filter(approved == 1)
baseline_app_rate <- mean(raw_data$approved)
baseline_pd <- mean(h_app$defaulted)
```

The historical baseline shows an **approval rate of `r fmt_pct(baseline_app_rate)`** and a **realized PD of `r fmt_pct(baseline_pd)`**. 

## Performance Stability: Vintage PD Analysis

Stability across time is the bedrock of credit policy. We observe the PD evolution across historical vintages.

```{r vintage-viz, echo=FALSE}
vintage_data <- h_app %>%
  group_by(vintage_month, legacy_decile) %>%
  summarize(PD = mean(defaulted), .groups = "drop")

ggplot(vintage_data, aes(x = vintage_month, y = PD, color = factor(legacy_decile))) +
  geom_line(size = 0.8, alpha = 0.7) +
  theme_strategic() +
  scale_y_continuous(labels = percent) +
  scale_color_viridis_d(option = "magma", end = 0.8) +
  labs(title = "Historical PD Stability per Decile", subtitle = "Vintage performance across the last 24 months.", x = "Vintage", y = "Observed PD", color = "Legacy Decile")
```

---

# 2. Strategic Frontiers & Stress Testing

We simulate the performance under macroeconomic stress (aggravation factors of 1.3, 1.5, and 1.8).

```{r frontiers}
# Calibrated Funnel for Realistic Results (~15-20% Approval)
build_stages <- function(scores, cutoffs = NULL) {
  stages <- list(
    stage_filter("kyc", "id_valid == TRUE & age >= 19"),
    # Operational leakage to keep results realistic
    stage_rate("fraud", base_rate = 0.95),
    stage_rate("manual_review", base_rate = 0.40),
    stage_rate("conversion", base_rate = 0.65)
  )
  if (!is.null(cutoffs)) {
    stages <- c(list(stage_cutoff("selection", cutoffs)), stages)
  }
  return(stages)
}

get_frontier <- function(scores, stress_factor) {
  p <- credit_policy(
    applicant_id_col = "id", 
    score_cols = scores, 
    current_approval_col = "approved", 
    actual_default_col = "defaulted", 
    risk_level_col = "legacy_decile",
    simulation_stages = build_stages(scores),
    stress_scenarios = list(stress_aggravation(stress_factor, by = "legacy_decile"))
  )
  res <- find_optimal_cutoffs(raw_data, p, cutoff_steps = 15, method = "analytical")
  analyze_tradeoffs(res)$pareto_frontier %>% mutate(Stress = stress_factor)
}

front_M_15 <- get_frontier(c("legacy_score", "challenger_score"), 1.5)
front_C_13 <- get_frontier("challenger_score", 1.3)
front_C_15 <- get_frontier("challenger_score", 1.5)
front_C_18 <- get_frontier("challenger_score", 1.8)

frontier_all <- bind_rows(
  front_C_13 %>% mutate(Label = "Challenger 1.3x"),
  front_C_15 %>% mutate(Label = "Challenger 1.5x"),
  front_C_18 %>% mutate(Label = "Challenger 1.8x"),
  front_M_15 %>% mutate(Label = "Matrix 1.5x")
)
```

```{r plot-frontier, echo=FALSE}
ggplot(frontier_all, aes(x = overall_approval_rate, y = overall_default_rate, color = Label)) +
  geom_line(size = 1.2) +
  geom_point(aes(x = baseline_app_rate, y = baseline_pd * 1.5), color = "black", size = 4, shape = 18) +
  annotate("text", x = baseline_app_rate + 0.01, y = baseline_pd * 1.5, label = "Legacy @ 1.5x", fontface = "bold", size = 3) +
  theme_strategic() +
  scale_x_continuous(labels = percent) +
  scale_y_continuous(labels = percent) +
  scale_color_manual(values = c("Challenger 1.3x" = "#F1C40F", "Challenger 1.5x" = "#E67E22", "Challenger 1.8x" = "#C0392B", "Matrix 1.5x" = "#27AE60")) +
  labs(title = "Efficiency Frontier Comparison", subtitle = "The Matrix Strategy dominates the replacement at the target approval rate.", x = "Approval Rate", y = "Portfolio PD (Expected)")
```

---

# 3. Strategic Profiling: The 4-Quadrant Matrix

We map applicant density across Legacy and Challenger deciles to identify optimization opportunities.

```{r matrix-viz, echo=FALSE}
iso_M_15 <- find_equivalent_policy(front_M_15, "approval_rate", baseline_app_rate)
m_cut_l <- if(nrow(iso_M_15)>0) as.numeric(iso_M_15$legacy_score[1]) else 450
m_cut_c <- if(nrow(iso_M_15)>0) as.numeric(iso_M_15$challenger_score[1]) else 500

matrix_agg <- raw_data %>%
  group_by(legacy_decile, challenger_decile) %>%
  summarize(Volume = n(), PD = mean(defaulted), .groups = "drop") %>%
  mutate(Participation = Volume / sum(Volume))

ggplot(matrix_agg, aes(x = factor(legacy_decile), y = factor(challenger_decile))) +
  geom_tile(aes(fill = Participation)) +
  scale_fill_gradient(low = "white", high = c_emerald, labels = percent) +
  geom_vline(xintercept = 4.5, linetype = "dashed", color = c_ruby) +
  geom_hline(yintercept = 5.5, linetype = "dashed", color = c_ruby) +
  annotate("text", x = 2, y = 8, label = "Swap-In", color = c_ruby, fontface = "bold") +
  annotate("text", x = 8, y = 2, label = "Swap-Out", color = c_ruby, fontface = "bold") +
  theme_strategic() +
  labs(title = "Strategic Concentration Matrix", x = "Legacy Decile", y = "Challenger Decile", fill = "Population %")
```

---

# 4. Master Risk Rating Table

Comparison of current strategy vs. proposed populations per institutional Risk Rating.

```{r risk-grouping, echo=FALSE}
# Final Simulations
p_matrix <- credit_policy("id", c("legacy_score", "challenger_score"), "approved", "defaulted", 
                         risk_level_col = "legacy_decile",
                         simulation_stages = build_stages(c("legacy_score", "challenger_score"), list(legacy_score=m_cut_l, challenger_score=m_cut_c)))
sim_matrix <- run_simulation(raw_data, p_matrix, method = "analytical", quiet = TRUE)

iso_C_15 <- find_equivalent_policy(front_C_15, "approval_rate", baseline_app_rate)
# Ensure at least one policy is found
if(nrow(iso_C_15) == 0) iso_C_15 <- data.frame(challenger_score = 550, overall_approval_rate=0.15, overall_default_rate=0.04)

p_chall <- credit_policy("id", "challenger_score", "approved", "defaulted", 
                        risk_level_col = "legacy_decile",
                        simulation_stages = build_stages("challenger_score", list(challenger_score=as.numeric(iso_C_15$challenger_score[1]))))
sim_chall <- run_simulation(raw_data, p_chall, method = "analytical", quiet = TRUE)

# Establish Risk Ratings
rg <- find_risk_groups(sim_matrix$data %>% filter(new_approval > 0), 
                      score_cols = c("legacy_decile", "challenger_decile"), 
                      default_col = "defaulted", time_col = "vintage", bins = 10)

# Deep Analysis Table
rated_matrix <- rg$data
# Joins using the generated bin names from find_risk_groups
rated_challenger <- sim_chall$data %>% filter(new_approval > 0) %>%
  rename(legacy_decile_bin = legacy_decile, challenger_decile_bin = challenger_decile) %>%
  left_join(rg$mapping, by = c("legacy_decile_bin", "challenger_decile_bin"))

risk_master <- rated_matrix %>%
  group_by(risk_rating) %>%
  summarize(
    `PD Obs Keep-In` = fmt_pct(mean(defaulted[scenario == "keep_in"], na.rm=TRUE)),
    `PD Exp Swap-In (M)` = fmt_pct(mean(simulated_default[scenario == "swap_in"], na.rm=TRUE)),
    `Vol Swap-In (M) %` = fmt_pct(sum(scenario == "swap_in") / n()), 
    .groups = "drop"
  ) %>%
  left_join(
    rated_challenger %>% group_by(risk_rating) %>% 
      summarize(`PD Exp Swap-In (C)` = fmt_pct(mean(simulated_default[scenario == "swap_in"], na.rm=TRUE)),
                `Vol Swap-In (C) %` = fmt_pct(sum(scenario == "swap_in") / n()), .groups = "drop"),
    by = "risk_rating"
  )

risk_master %>%
  rename(Rating = risk_rating) %>%
  kbl(caption = "Performance Deep Dive: Swap-In Analysis per Risk Rating") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## Longitudinal Stability per Risk Rating

We verify that the generated risk groups maintain their hierarchical separation and stability across time cohorts.

```{r stability-plot, echo=FALSE}
plot(rg) + theme_strategic() + labs(title = "Risk Rating Stability", subtitle = "Monotonic PD ranking across vintages.")
```

---

# Strategic Decision Verdict

The **Matrix Strategy** is the recommended path. It filters "Swap-Out" risks that a pure replacement model would accept, achieving a portfolio PD **`r fmt_pct(as.numeric(iso_C_15$overall_default_rate[1]) - as.numeric(iso_M_15$overall_default_rate[1]))`** lower than a standalone Challenger transition at parity approval.

### Recommendation
1.  **Migrate** to hybrid cutoffs: Legacy >= `r m_cut_l` & Challenger >= `r m_cut_c`.
2.  **Monitor** Risk Ratings 1-7 for any macroeconomic drift.

---
**Prepared with creditools v0.4.0**  
*Strategic Credit Policy Proposal*
