---
title: "Multi-Stage Credit Funnel Optimization"
subtitle: "Strategic Transition: Identity, Age, Bureau and Score Matrix Strategy"
author: "Matheus Pasche (Strategic Analyst Portfolio)"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    css: !expr "system.file('rmarkdown/templates/html_vignette/resources/vignette.css', package = 'rmarkdown')"
vignette: >
  %\VignetteIndexEntry{Multi-Stage Credit Funnel Optimization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 8, fig.height = 4.5,
  out.width = "100%",
  message = FALSE, warning = FALSE,
  cache = FALSE,
  dpi = 130
)
```

```{r library}
library(creditools)
library(dplyr)
library(ggplot2)
library(tidyr)
```

# Introduction

In modern credit architectures, decisioning is rarely a single "stop/go" signal. It is a **sequential funnel** designed to filter out risks at various layers: demographic validation, regulatory constraints, bureau analysis, and finally, statistical score optimization.

This case study demonstrates the use of the `creditools` package to simulate and optimize a **Complex Multi-Stage Funnel** using the built-in `applicants` dataset.

# 1. Scenario Setup: The Built-in Portfolio

We will use the package's built-in `applicants` dataset, containing 20,000 synthetic profiles with complex demographics.

```{r data}
# Load the built-in dataset
data(applicants)

# View a summary of the complex demographics
applicants %>%
  select(age, bureau_derogatory, id_valid) %>%
  summary()
```

# 2. Sequential Funnel Design

We will move from a simple legacy strategy to a multi-stage funnel:

1.  **Stage 1: Identity Filter** (Proxy for KYC validation).
2.  **Stage 2: Age Constraint** (Regulatory/Policy constraint).
3.  **Stage 3: Bureau Hard-Cut** (Negative registries check).
4.  **Stage 4: Optimized Score Matrix** (Balancing Approval vs Default).

## The Legacy Baseline
Historically, the policy only considered the `old_score > 600`.

```{r legacy-policy}
legacy_policy <- credit_policy(
  applicant_id_col = "id",
  score_cols = "old_score",
  current_approval_col = "approved",
  actual_default_col = "defaulted"
) %>%
  add_stage(stage_cutoff(name = "legacy_cutoff", cutoffs = list(old_score = 600)))

# Run legacy simulation
legacy_results <- run_simulation(applicants, legacy_policy)
```

## Designing the Optimized Funnel
We first find the optimal cutoffs for our scores, *assuming* the previous demographic filters are already in place.

```{r optimization}
# Define the demographic filters base
base_funnel <- credit_policy(
  applicant_id_col = "id",
  score_cols = c("old_score", "new_score"),
  current_approval_col = "approved",
  actual_default_col = "defaulted"
) %>%
  add_stage(stage_filter(name = "id_check", condition = "id_valid == 1")) %>%
  add_stage(stage_filter(name = "age_check", condition = "age >= 18")) %>%
  add_stage(stage_filter(name = "bureau_check", condition = "bureau_derogatory < 500"))

# Optimize the final stage (Score Matrix)
# We want to maximize approval while keeping global default rate < 4%
opt_results <- find_optimal_cutoffs(
  data = applicants,
  config = base_funnel,
  cutoff_steps = 15,
  target_default_rate = 0.04,
  min_approval_rate = 0.20,
  method = "analytical"
)

# Best combination found
print(opt_results)
```

# 3. Decision Frontier: The Risk-Reward Landscape

```{r frontiers}
# Analyze the Pareto frontier of all tested cutoff combinations
tradeoffs <- analyze_tradeoffs(opt_results)

# Visualize
visualize_tradeoffs(tradeoffs, type = "pareto") +
  labs(title = "Decision Frontier: Complex Funnel + Matrix Optimization")
```

# 4. Stress Testing: Market Resilience

We now assemble the full pipeline and test it under a 30% default aggravation scenario (simulating a severe economic downturn).

```{r stress-testing}
# Assemble final optimized policy
optimized_funnel <- base_funnel %>%
  add_stage(stage_cutoff(
    name = "score_matrix", 
    cutoffs = list(
      old_score = opt_results$old_score,
      new_score = opt_results$new_score
    )
  ))

# Stress test the funnel
stress_results <- run_tradeoff_analysis(
  data = applicants,
  base_policy = optimized_funnel,
  vary_params = list(
    aggravation_factor = c(1.0, 1.3)
  )
)

stress_results
```

# 5. Comparative Performance

```{r comparison}
# Helper to extract metrics
get_sum <- function(res, name) {
  tibble(
    Strategy = name,
    Approval_Rate = mean(res$data$new_approval),
    Default_Rate = sum(res$data$new_approval * res$data$simulated_default, na.rm=T) / 
                   sum(res$data$new_approval)
  )
}

legacy_sum <- get_sum(legacy_results, "Legacy (Score > 600)")

# Calculate final funnel performance (Baseline scenario)
final_res <- run_simulation(applicants, optimized_funnel, quiet = TRUE)
optimized_sum <- get_sum(final_res, "Optimized Multi-Stage Funnel")

bind_rows(legacy_sum, optimized_sum)
```

# Conclusion

By leveraging a multi-stage funnel and automated optimization, we successfully:
1.  **Introduced** mandatory demographic and bureau safety filters.
2.  **Optimized** the final scoring stage to meet precise risk targets.
3.  **Increased** overall portfolio transparency by simulating the impact of each stage.

This approach ensures that credit policies are not just predictive, but strategically aligned with business constraints and resilient to market stress.
