---
title: "Trade-off Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trade-off Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates how to use `creditools` to perform a trade-off analysis between approval rates and default rates under different stress scenarios and credit score cutoffs.

## 1. Setup

First, we load the necessary packages and `creditools`.

```{r setup}
# Ensure you have these packages installed:
# install.packages(c("devtools", "tidyverse", "cli", "ggplot2", "furrr", "fs"))
devtools::load_all()
library(tidyverse)
library(ggplot2)
library(scales)
```

## 2. Analysis Parameters

We define the parameters for our simulation. For the purpose of this vignette, we will use a smaller dataset to keep the execution time low.

```{r params-vignette}
N_APPLICANTS <- 20000 # Larger sample for more stable results
SEED <- 42
CUTOFF_STEP <- 50
AGGRAVATION_FACTORS <- c(1.2, 1.5)
```

## 3. Data Generation

We generate a sample dataset using `generate_sample_data`.

```{r data}
analytical_base <- creditools::generate_sample_data(
  n_applicants = N_APPLICANTS,
  seed = SEED
)
# Pre-calculate deciles needed for stratified aggravation
analytical_base$new_score_decile <- dplyr::ntile(analytical_base$new_score, 10)
```

## 4. Simulation using the Wrapper Function

Here, we use the powerful `run_tradeoff_analysis` function. We define a `base_policy` and then specify which parameters we want to vary in `vary_params`. The function handles creating the parameter grid and running all simulations.

```{r simulation}
# Define a base policy
base_policy <- credit_policy(
  applicant_id_col = "id",
  score_cols = c("old_score", "new_score"),
  current_approval_col = "approved",
  actual_default_col = "defaulted",
  risk_level_col = "new_score_decile"
)

# Define parameters to vary
min_score <- floor(min(analytical_base$new_score) / CUTOFF_STEP) * CUTOFF_STEP
max_score <- ceiling(max(analytical_base$new_score) / CUTOFF_STEP) * CUTOFF_STEP
cutoff_points <- seq(min_score, max_score, by = CUTOFF_STEP)

vary_params <- list(
  new_score_cutoff = cutoff_points,
  aggravation_factor = AGGRAVATION_FACTORS
)

# Run the tradeoff analysis
# For a real analysis, you could set parallel = TRUE after setting up a future plan
# e.g. future::plan(future::multisession)
simulation_outputs <- run_tradeoff_analysis(
  data = analytical_base,
  base_policy = base_policy,
  vary_params = vary_params,
  parallel = FALSE
)
```

## 5. Visualization

We visualize the trade-off between approval rate and default rate. This plot clearly shows how, for a given approval rate, the default rate increases under more stressful scenarios.

```{r plot}
plot_data <- simulation_outputs %>%
  mutate(stress_scenario = paste0(round((aggravation_factor - 1) * 100), "% Aggravation"))

tradeoff_plot <- ggplot(plot_data, aes(x = approval_rate, y = default_rate, color = stress_scenario)) +
  geom_line(size = 1.2) +
  geom_point(size = 2.5) +
  labs(
    title = "Trade-off: Approval Rate vs. Default Rate",
    subtitle = "Analysis of score cutoffs under different stress scenarios for swap-ins",
    x = "Overall Approval Rate",
    y = "Average Default Rate (on Approved)",
    color = "Stress Scenario"
  ) +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )

print(tradeoff_plot)
```

## 6. Deeper Analysis

Now we can use the results to answer more specific business questions.

### Where do default rates equalize?

A common question is: at what cutoff point does the default rate of a less aggressive policy (e.g., 1.2x aggravation) equal the default rate of a more aggressive one? This can inform decisions about risk appetite.

Since the aggravation factor only affects the `swap_in` population, a higher cutoff means a smaller, potentially riskier `swap_in` pool, causing the default rate lines to converge.

```{r intersect-plot}
# Plot default rate vs. cutoff to see the intersection
ggplot(plot_data, aes(x = new_score_cutoff, y = default_rate, color = stress_scenario)) +
  geom_line(size = 1.2) +
  geom_point(size = 2.5) +
  labs(
    title = "Default Rate by Cutoff and Stress Scenario",
    subtitle = "Finding the point where default rates converge",
    x = "New Score Cutoff",
    y = "Average Default Rate (on Approved)",
    color = "Stress Scenario"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")

# Find the intersection point programmatically
intersection_data <- plot_data %>%
  select(new_score_cutoff, stress_scenario, default_rate) %>%
  pivot_wider(names_from = stress_scenario, values_from = default_rate)

# Find the cutoff where the difference changes sign
# Note: Column names are dynamic based on AGGRAVATION_FACTORS
col_names <- paste0(round((AGGRAVATION_FACTORS - 1) * 100), "% Aggravation")
intersection_data <- intersection_data %>%
  mutate(diff = .data[[col_names[1]]] - .data[[col_names[2]]])

# Use linear approximation to find the intersection
# Get the two points where the sign of the difference flips
cross_points <- intersection_data %>%
  filter(diff > 0 | lag(diff < 0, default = TRUE)) %>%
  slice(1:2)

if (nrow(cross_points) == 2) {
  intersection_cutoff <- approx(x = cross_points$diff, y = cross_points$new_score_cutoff, xout = 0)$y
  cat(paste("\nEstimated cutoff where default rates equalize:", round(intersection_cutoff)))
}
```

### Risk Profile at the Intersection Point

What does the risk profile of our approved population look like at this equalizing cutoff point? We can run a single simulation there and analyze the distribution of risk deciles.

```{r decile-dist-plot}
# Use the found cutoff, or a representative one if none was found
analysis_cutoff <- if (exists("intersection_cutoff") && !is.na(intersection_cutoff)) round(intersection_cutoff) else 600

# Rerun one simulation at this specific cutoff
sim_at_cutoff <- run_simulation(
  data = analytical_base,
  policy = credit_policy(
    applicant_id_col = "id",
    score_cols = c("old_score", "new_score"),
    current_approval_col = "approved",
    actual_default_col = "defaulted",
    risk_level_col = "new_score_decile",
    simulation_stages = list(
      stage_cutoff(name = "credit", cutoffs = list(new_score = analysis_cutoff))
    ),
    # We don't need stress scenarios for this part
    stress_scenarios = list()
  )
)

# Compare decile distributions of keep-ins vs swap-ins
decile_dist <- sim_at_cutoff$data %>%
  filter(scenario %in% c("keep_in", "swap_in")) %>%
  count(scenario, new_score_decile) %>%
  group_by(scenario) %>%
  mutate(proportion = n / sum(n))

ggplot(decile_dist, aes(x = factor(new_score_decile), y = proportion, fill = scenario)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = paste("Risk Profile Comparison at Cutoff =", analysis_cutoff),
    subtitle = "Distribution of risk deciles for Keep-ins vs. Swap-ins",
    x = "New Score Decile",
    y = "Proportion of Population",
    fill = "Scenario"
  ) +
  theme_minimal(base_size = 14)
```

This plot shows, as expected, that the `swap_in` population is concentrated in the lower (riskier) deciles compared to the `keep_in` population. This is the portfolio segment whose risk we are simulating.
