---
title: "Trade-off Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trade-off Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates how to use `creditools` to perform a trade-off analysis between approval rates and default rates under different stress scenarios and credit score cutoffs.

## 1. Setup

First, we load the necessary packages and `creditools`.

```{r setup}
library(creditools)
library(tidyverse)
library(ggplot2)
library(scales)
```

## 2. Analysis Parameters

We will use the package's built-in `applicants` dataset, which contains 20,000 synthetic profiles.

```{r data}
# Load built-in data
data(applicants)

# Parameters for stress evaluation
CUTOFF_STEP <- 50
AGGRAVATION_FACTORS <- c(1.2, 1.5)

# Pre-calculate deciles needed for stratified aggravation
applicants_prepared <- applicants %>%
  mutate(new_score_decile = dplyr::ntile(new_score, 10))
```

## 3. Simulation using the Wrapper Function

Here, we use the powerful `run_tradeoff_analysis` function. We define a `base_policy` and then specify which parameters we want to vary in `vary_params`.

```{r simulation}
# Define a base policy
base_policy <- credit_policy(
  applicant_id_col = "id",
  score_cols = c("old_score", "new_score"),
  current_approval_col = "approved",
  actual_default_col = "defaulted",
  risk_level_col = "new_score_decile"
)

# Define parameters to vary
min_score <- floor(min(applicants_prepared$new_score) / CUTOFF_STEP) * CUTOFF_STEP
max_score <- ceiling(max(applicants_prepared$new_score) / CUTOFF_STEP) * CUTOFF_STEP
cutoff_points <- seq(min_score, max_score, by = CUTOFF_STEP)

vary_params <- list(
  new_score_cutoff = cutoff_points,
  aggravation_factor = AGGRAVATION_FACTORS
)

# Run the tradeoff analysis
simulation_outputs <- run_tradeoff_analysis(
  data = applicants_prepared,
  base_policy = base_policy,
  vary_params = vary_params,
  parallel = FALSE
)
```

## 4. Visualization with Baseline Comparison

Now, we'll calculate the performance of the "current" policy (based on `old_score`) to serve as a baseline in our analysis.

```{r baseline-analysis}
# --- Baseline Analysis for Old Score ---
old_score_cutoffs <- seq(floor(min(applicants_prepared$old_score)), ceiling(max(applicants_prepared$old_score)), by = 10)

baseline_results <- purrr::map_dfr(old_score_cutoffs, function(cutoff) {
    approved_population <- applicants_prepared %>%
        dplyr::filter(old_score >= cutoff)
    
    approval_rate <- if (nrow(applicants_prepared) > 0) nrow(approved_population) / nrow(applicants_prepared) else 0
    
    hired_approved_population <- approved_population %>% dplyr::filter(hired == 1)
    default_rate <- if (nrow(hired_approved_population) > 0) {
        mean(hired_approved_population$defaulted, na.rm = TRUE)
    } else {
        0
    }
    
    tibble::tibble(
        cutoff = cutoff,
        approval_rate = approval_rate,
        default_rate = default_rate
    )
})
```

We visualize the trade-off between approval rate and default rate.

```{r plot}
plot_data <- simulation_outputs %>%
  dplyr::mutate(stress_scenario = paste0(round((aggravation_factor - 1) * 100), "% Aggravation"))

ggplot(plot_data, aes(x = approval_rate, y = default_rate, color = stress_scenario)) +
  geom_line(size = 1.2) +
  geom_point(size = 2.5) +
  # Add the baseline for the old score
  geom_line(
    data = baseline_results,
    aes(x = approval_rate, y = default_rate),
    color = "black", linetype = "dashed", size = 1.1, inherit.aes = FALSE
  ) +
  labs(
    title = "Trade-off: Approval Rate vs. Default Rate",
    subtitle = "Comparing New Score Policy vs. Current Policy Baseline (Dashed)",
    x = "Overall Approval Rate",
    y = "Average Default Rate (on Approved)",
    color = "New Policy Stress Scenario"
  ) +
  scale_x_continuous(labels = scales::percent_format()) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )
```

## 5. Conclusion

By using built-in datasets and standardized analysis pipelines, `creditools` enables rapid assessment of score performance across various economic scenarios.
